### 2021-11-16 manuscript/hiv-gwas.tex

\begin{flushleft}
	Sarah Nadeau\textsuperscript{1,2}, Christian Thorball\textsuperscript{2,3}, Roger Kouyos\textsuperscript{4,5}, Huldrych F. Günthard\textsuperscript{4,5}, Jürg Böni\textsuperscript{4}, Sabine Yerly\textsuperscript{6}, Matthieu Perreau\textsuperscript{7}, Thomas Klimkait\textsuperscript{8}, Andri Rauch\textsuperscript{9}, Manuel Battegay\textsuperscript{10}, Matthias Cavassini\textsuperscript{11}, Alexandra Calmy\textsuperscript{12}, Pietro Vernazza\textsuperscript{13}, Enos Bernasconi\textsuperscript{14}, Jacques Fellay\textsuperscript{2,3}, Venelin Mitov\textsuperscript{1,2,*}, Tanja Stadler\textsuperscript{1,2,*}, and the Swiss HIV Cohort Study (SHCS)
\end{flushleft}

\noindent
\textsuperscript{1}\emph{Department of Biosystems Science and Engineering, ETH Zürich, Basel} \\
\textsuperscript{2}\emph{Swiss Institute of Bioinformatics, Lausanne} \\
\textsuperscript{3}\emph{Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne} \\
\textsuperscript{4}\emph{Institute of Medical Virology, University of Zürich, Zürich} \\
\textsuperscript{5}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zürich, University of Zürich, Zürich} \\
\textsuperscript{6}\emph{Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva} \\
\textsuperscript{7}\emph{Division of Immunology and Allergy, University Hospital Lausanne, Lausanne} \\
\textsuperscript{8}\emph{Molecular Virology, Department of Biomedicine - Petersplatz, University of Basel, Basel} \\
\textsuperscript{9}\emph{Department of Infectious Diseases, Berne University Hospital and University of Berne, Berne} \\
\textsuperscript{10}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel} \\
\textsuperscript{11}\emph{Division of Infectious Diseases, University Hospital Lausanne, Lausanne} \\
\textsuperscript{12}\emph{HIV/AIDS Unit, Infectious Disease Service, Geneva University Hospital, Geneva} \\
\textsuperscript{13}\emph{Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen} \\
\textsuperscript{14}\emph{Division of Infectious Diseases, Regional Hospital Lugano, Lugano} \\
\textsuperscript{*}\emph{Co-last authors}

@@

\begin{abstract}
	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all effect disease traits. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits. However, GWAS for host genetic determinants of infectious disease traits do not account for potentially confounding or extraneous pathogen effects that are heritable from one infection partner to another in transmission chains.  We propose a novel method to consider the full genome of each patient's infecting strain, remove strain-specific effects on a trait, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable. When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association with spVL in the \emph{CCR5} and major histocompatability complex (MHC) gene regions compared to a standard GWAS. Our simulation study confirms that estimated heritability and selection parameters for HIV-1 mean standard GWAS are quite robust to pathogen effects. Our framework may improve GWAS for other diseases where pathogen effects are stronger and more correlated amongst individuals in a cohort.
\end{abstract}

@@

% What is GWAS and why is it important
A key goal of genome-wide association studies (GWAS) is to understand the genetic basis of phenotypic variation among individuals. In a typical GWAS, millions of genetic variants from across the human genome are screened for statistical association with a trait of interest. Ideally, this procedure identifies variants that are located in or are in linkage disequilibrium with alleles that directly effect the trait. If GWAS finds a variant strongly associated with a disease trait, the gene product may be a good drug target \citep{Okada2014}. Even if no single variant has a strong association, many small associations can be aggregated into a polygenic risk score to identify high-risk individuals \citep{Dudbridge2013}.

@@

% How heritable is spVL (human side)
Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that display viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed. CCR5 proteins are a signal to the virus that help it find the right cells to infect.

% How heritable is spVL (viral side)
In addition to these human genetic factors, is also well-recognized that viral genetic factors effect spVL. As mentioned, heritability from the viral side is typically measured using broad-sense heritability. Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty). Estimates using phylogenetic methods on large UK and Swiss cohorts by \citep{Mitov2018} and \citep{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%. However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016}. If the virus can change it's ``puzzle piece'' shape to escape MHC proteins, infected cells can go undetected. This means that MHC variants effect spVL largely via selection on the virus \citep{Bartha2017}. In summary, human genetic factors play a role in determining spVL, but these effects may be mediated by viral genetic variation.

% The gap
Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL. In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors. This has several potential negative consequences: (i) viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, (ii) variability due to viral effects would make it more challenging to identify human variants of small effect, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.

% Overview of relevant methods and why we don't use them
Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry in a cohort can give rise to spurious genetic correlations with a trait. Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}. More recently, \citep{Power2016} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS. Two state-of-the-art methods exist for this \citep{Earle2016, collins_phylogenetic_nodate}. 
% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework. \citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry. 
However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait. In contrast, we want to estimate effects of $human$ genetic variants on a trait, taking into account pathogen effects. \citep{Naret2018} developed a relevant method for this task. The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing. This should correct for trait correlations due to shared pathogen ancestry. However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).

@@

\begin{proof}
	\begin{align}\label{eq:pdfG}
	\begin{split}
		f\big(\bm{g}_v,\ \bm{e}\big) &= f\big(\bm{g}_v|\ \bm{e}\big) \times f\big(\bm{e}\big) \\
	&= f\big(\bm{g}_v\big) \times f\big(\bm{e}\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{e};\ \bm{0}, \mathbf{\Sigma}_E\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g}_v;\ \bm{0}, \mathbf{\Sigma}_E\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_e\big)
	\end{split}
\end{align}

@@

Next, we characterized the evolutionary scenarios under which our framework can actually improve GWAS power. We used the true positive rate (TPR) to evaluate the fraction of simulated host genetic variants with direct effects we could recover as being significantly associated with the trait. We performed three different GWAS for each simulated dataset: the first represents an ideal in which we can exactly know and remove pathogen effects from trait values, the second is using our method to estimate this value and remove it, and the third represents a standard GWAS using the scaled trait value. Figure \ref{fig:simulationResults}B shows that our framework can improve the TPR in simulated scenarios where selection strength $<$ 10 time$^{-1}$ and heritability $>$ 45\%. If were were able to perfectly estimate and remove pathogen effects on a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable. We estimate approximately 25\% to be the heritability threshold above which GWAS power is negatively impacted by pathogen effects. In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by estimating and removing pathogen effects using information from the pathogen phylogeny.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.75\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows that our method (left) can generate more accurate estimates of host genetic effects than the trait value, scaled by its mean (right). (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The black point represents our estimates for the heritability and selection strength of spVL based on Swiss data. RMSE = Root mean square error, TPR = True positive rate.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

@@

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. In the ``GWAS with standard trait value'' we used the unmodified trait value, our calculated spVL values, as the GWAS response variable. In the ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (figure \ref{fig:qq-plots}). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (figure \ref{tab:gwas-top-associations}). Finally table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}. In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

@@

% Contextualize the study & address limitations
Our method relies on the POUMM, a model of continuous trait evolution that accounts for heritable and non-heritable effects on a trait, as well as selection. Using this model, we estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the Swiss HIV Cohort Study. Compared to previous studies, this estimate is at the higher end \citep{Mitov2018}. Also using the POUMM, \citep{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citep{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort. We note that our sample size (N = 1493 individuals) is quite a bit smaller than these other studies. This might be because we restricted samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases. Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream GWAS methods. So although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.

Instead, the main novelty of our approach was to correct the spVL trait prior to GWAS by estimating and removing pathogen effects. In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced. Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny. Unfortunately, given our estimates for the heritability of spVL and the selection strength on this trait, our simulation results show that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (figure \ref{fig:simulationResults}). Our method is anticipated to be more useful for disease traits that are highly heritable from infection partner to infection partner and  maintain a high correlation between sampled individuals. In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations. So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to be meaningful.

Our method is convenient for GWAS because it is simply a pre-processing step that produces an alternate response variable for GWAS association tests. Therefore, it is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}). The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project github at TODO.

@@

For GWAS, we retained data from 1,392 of the 1,493 individuals with European ancestry and not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}). These were 227 females and 1165 males. Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure \ref{fig:PCA}). Kinship was evaluated using PLINK version 2.00a2LM \citep{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375. Initial host genotyping quality control and imputation were done by the SHCS. Subsequent genotyping quality control was performed using PLINK version 2.3 \citep{Chang2015}. We used the options --maf 0.01, --geno 0.01, and --hwe 0.00005 to remove variants with minor allele frequency less than 0.01, missing call rate greater than 0.05, or Hardy-Weinberg equilibrium exact test p-value less than 5x10$^-5$. After quality filtering, approximately 6.2 million genetic variants from the 1,392 individuals were retained for GWAS (Table \ref{tab:variant-filtering}).

@@

This study was funded by ETH Zurich. We would like to thank Esposito Mauro and Alexandra Scherrer from the SHCS data center, who were especially helpful in retrieving data, and Michael Landis, who shared a LaTeX template for graphical model drawing. Finally and most importantly, we thank the participants of the Swiss HIV Cohort Study for contributing their samples and making this research possible.

>>>>

\begin{flushleft}
	Sarah Nadeau\textsuperscript{1,2}, Christian W. Thorball\textsuperscript{3}, Roger Kouyos\textsuperscript{4,5}, Huldrych F. Günthard\textsuperscript{4,5}, Jürg Böni\textsuperscript{4}, Sabine Yerly\textsuperscript{6}, Matthieu Perreau\textsuperscript{7}, Thomas Klimkait\textsuperscript{8}, Andri Rauch\textsuperscript{9}, Hans H. Hirsch\textsuperscript{8,10,11}, Matthias Cavassini\textsuperscript{12}, Pietro Vernazza\textsuperscript{13}, Enos Bernasconi\textsuperscript{14}, Jacques Fellay\textsuperscript{2,3,15}, Venelin Mitov\textsuperscript{1,2,*}, Tanja Stadler\textsuperscript{1,2,*}, and the Swiss HIV Cohort Study (SHCS)
\end{flushleft}

\noindent
\textsuperscript{1}\emph{Department of Biosystems Science and Engineering, ETH Zürich, Basel} \\
\textsuperscript{2}\emph{Swiss Institute of Bioinformatics, Lausanne} \\
\textsuperscript{3}\emph{Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland} \\
\textsuperscript{4}\emph{Institute of Medical Virology, University of Zürich, Zürich} \\
\textsuperscript{5}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zürich, University of Zürich, Zürich} \\
\textsuperscript{6}\emph{Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva} \\
\textsuperscript{7}\emph{Division of Immunology and Allergy, University Hospital Lausanne, Lausanne} \\
\textsuperscript{8}\emph{Department of Biomedicine, University of Basel, Basel} \\
\textsuperscript{9}\emph{Department of Infectious Diseases, Berne University Hospital and University of Berne, Berne} \\
\textsuperscript{10}\emph{Division of Clinical Virology, University Hospital Basel, Basel} \\
\textsuperscript{11}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel} \\
\textsuperscript{12}\emph{Division of Infectious Diseases, University Hospital Lausanne, Lausanne} \\
\textsuperscript{13}\emph{Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen} \\
\textsuperscript{14}\emph{Division of Infectious Diseases, Regional Hospital Lugano, Lugano} \\
\textsuperscript{15}\emph{Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne} \\
\textsuperscript{*}\emph{Co-last authors}

@@

\begin{abstract}
	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all effect disease traits. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits. However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.  We propose a novel method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits. In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations. When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatability complex (MHC) gene regions compared to standard GWAS. Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects. Our framework may improve GWAS for other diseases where pathogen effects are stronger and more phylogenetically correlated amongst individuals in a cohort.
\end{abstract}

@@

% What is GWAS and why is it important
A key goal of genome-wide association studies (GWAS) is to understand the genetic basis of phenotypic variation among individuals. In a typical GWAS, millions of genetic variants from across the human genome are screened for statistical association with a trait of interest. Ideally, this procedure identifies variants that are located in, or are in linkage disequilibrium with, alleles that directly affect the trait. If GWAS finds a variant strongly associated with a disease trait, the gene product may be a good drug target \citep{Okada2014}. Even if no single variant has a strong association, many small associations can be aggregated into a polygenic risk score to identify high-risk individuals \citep{Dudbridge2013}.

@@

% How heritable is spVL (human side)
Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest study to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed. CCR5 proteins help the virus infect target cells.

% How heritable is spVL (viral side)
In addition to these human genetic factors, it is well-recognized that viral genetic factors effect spVL. As mentioned, heritability from the viral side is typically measured using broad-sense heritability. Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty). Estimates using phylogenetic methods on large UK and Swiss cohorts by \citep{Mitov2018} and \citep{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%. However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021}. If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected. This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}. In summary, human genetic factors play a role in determining spVL, but these effects may be mediated by viral genetic variation.

% The gap
Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL. In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors. This has several potential negative consequences, namely that (i) viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, (ii) variability due to viral effects would make it more challenging to identify human variants of small effect, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.

% Overview of relevant methods and why we don't use them
Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry, especially between close relatives, can give rise to spurious genetic correlations with a trait. Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}. More recently, \citep{Power2016} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS. Two state-of-the-art methods exist for this \citep{Earle2016, collins_phylogenetic_nodate}. 
% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework. \citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry. 
However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait. In contrast, we want to estimate effects of $human$ genetic variants on a trait, taking into account pathogen effects. \citep{Naret2018} developed a relevant method for this task. The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing. This should correct for trait correlations due to shared pathogen ancestry. However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).

@@

\begin{proof}
	\begin{align}\label{eq:pdfG}
	\begin{split}
		f\big(\bm{g}_v,\ \bm{e}\big) &= f\big(\bm{g}_v|\ \bm{e}\big) \times f\big(\bm{e}\big) \\
	&= f\big(\bm{g}_v\big) \times f\big(\bm{e}\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{e};\ \bm{0}, \mathbf{\Sigma}_E\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g}_v;\ \bm{0}, \mathbf{\Sigma}_E\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_E\big)
	\end{split}
\end{align}

@@

Next, we characterized the evolutionary scenarios under which our framework can actually improve GWAS power. We used the true positive rate (TPR) to evaluate the fraction of simulated host genetic variants with direct effects we could recover as being significantly associated with the trait. We performed three different GWAS for each simulated dataset: the first represents an ideal in which we can exactly know and remove pathogen effects from trait values, the second is using our method to estimate this value and remove it, and the third represents a standard GWAS using the scaled trait value. Figure \ref{fig:simulationResults}B shows that our framework can improve the TPR in simulated scenarios where selection strength $<$ 10 time$^{-1}$ and heritability $>$ 45\%. If we were able to perfectly estimate and remove pathogen effects on a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable. We estimate approximately 25\% to be the heritability threshold above which GWAS power is negatively impacted by pathogen effects. In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by estimating and removing pathogen effects using information from the pathogen phylogeny.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.75\linewidth]{figures/simulation_results}
		\caption{Results from the simulation study. We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters. For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values. (A) shows that our method (left) can generate more accurate estimates of host genetic effects than the trait value, scaled by its mean (right). (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right). Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples. The black point represents our estimates for the heritability and selection strength of spVL based on Swiss HIV data. RMSE = Root mean square error, TPR = True positive rate.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

@@

We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes. In the ``GWAS with standard trait value'' we used the unmodified trait value, our calculated spVL values, as the GWAS response variable. In the ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL. Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS. Q-Q plots show the distribution of p-values are very similar as well (figure \ref{fig:qq-plots}). Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions. Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region. Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$. However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (figure \ref{tab:gwas-top-associations}). Finally table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}. In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.

@@

% Contextualize the study & address limitations
Our method relies on the POUMM, a model of continuous trait evolution that accounts for heritable and non-heritable effects on a trait, as well as selection. Using this model, we estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the Swiss HIV Cohort Study. Compared to previous studies, this estimate is at the higher end \citep{Mitov2018}. Also using the POUMM, \citep{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citep{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort. We note that our sample size (N = 1493 individuals) is quite a bit smaller than these other studies. This might be because we restricted samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases. Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream GWAS methods. Although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.

Instead, the main novelty of our approach was to correct the spVL trait prior to performing the GWAS, thereby estimating and removing pathogen effects. In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced. Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny.

Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals. In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations. So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.

Given our estimates for the heritability of spVL and the selection strength on this trait using Swiss HIV cohort data, our simulation results reveal that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (figure \ref{fig:simulationResults}). Indeed the results of our  GWAS method presented here and a standard GWAS are very similar. Future applications might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g. \cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.

Our method is convenient for GWAS because it is simply a pre-processing step that produces an alternate response variable for GWAS association tests. Therefore, it is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}). The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS.

@@

For GWAS, we retained data from 1,392 of the 1,493 individuals with European ancestry and not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}). These were 227 females and 1165 males. Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure \ref{fig:PCA}). Kinship was evaluated using PLINK version 2.3 \citep{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375. Initial host genotyping quality control and imputation were done as in \cite{Thorball2021GeneticLymphoma}. Subsequent genotyping quality control was performed using PLINK version 2.3 \citep{Chang2015}. We used the options --maf 0.01, --geno 0.01, and --hwe 0.00005 to remove variants with minor allele frequency less than 0.01, missing call rate greater than 0.05, or Hardy-Weinberg equilibrium exact test p-value less than 5x10$^-5$. After quality filtering, approximately 6.2 million genetic variants from the 1,392 individuals were retained for GWAS (Table \ref{tab:variant-filtering}).

@@

This study was funded by ETH Zurich. We would like to thank Esposito Mauro and Alexandra Scherrer from the SHCS data center, who were especially helpful in retrieving data, Alexandra Calmy and her team from the Geneva University Hospital for data contributions, and Michael Landis, who shared a LaTeX template for graphical model drawing. Finally and most importantly, we thank the participants of the Swiss HIV Cohort Study for contributing their samples and making this research possible.
